---
title: 'Shiny Application and Reproducible Pitch'
subtitle: 'Analysing the COVID-19 pandemic and mapping its spread among CARICOM Member States'
author: "Yohance Nicholas"
date: "May 5th 2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

This Reproducible Pitch was created in partial fulfilment of the [Developing Data Products Course](https://www.coursera.org/learn/data-products/home/welcome) which comprises one of the five courses necessary for the Data Science: Statistics and Machine Learning Specialization offered by Johns Hopikins University through Coursera. 

This assignment challenged candidates to Create a reproducible pitch to sell the shiny app which was also developed as part of this coursework assignment. Candidates get 5 slides (inclusive of the title slide) to pitch a your app. Once completed, candidates must create a web page using Slidify or Rstudio Presenter with an html5 slide deck.

The presentation must fulfill the following requirements:

- 5 slides to pitch our idea done in Slidify or Rstudio Presenter
- Your presentation pushed to github or Rpubs
- A link to your github or Rpubs presentation pasted into the provided text box

All other coursework projects completed as part of this course can be found at my [GitHub repository](https://yohance-nicholas.github.io/Developing-Data-Products/) for this course. 

## Data Sources

With a view to map the spread of the disease thus far, I have elected to use two main data sources. Firstly, to obtain the most current data on the incidence of COVID-19, I have opted to utilise the data colelcted by the [Johns Hopkins Coronavirus Resource Centre](https://coronavirus.jhu.edu/). The [2019 Novel Coronavirus COVID-19 (2019-nCoV) Data Repository by Johns Hopkins CSSE](https://github.com/CSSEGISandData/COVID-19) is compiled from a cross section of sources daily. 

To supplement this data with relevant socio-demographic data, I have opted to utilise the [World Development Indicator Database](http://datatopics.worldbank.org/world-development-indicators/) maintained by the World Bank Group. The World Development Indicators is a compilation of relevant, high-quality, and internationally comparable statistics about global development and the fight against poverty. The database contains 1,600 time series indicators for 217 economies and more than 40 country groups, with data for many indicators going back more than 50 years.

A data cleaning script was prepared to pull the most current data on a range of issues. The script can be found in the [GitHub repository](https://yohance-nicholas.github.io/Developing-Data-Products/) created to store the content and code generated in the completion of this course.

## Overview and Essential Links

### Overview

This app allows for rudimentary exploratory data analysis of the COVID-19 pandemic database. The app depicts the bivariate relationships between variables and presents Least Square Regression models to assess the strength of the relationship.

### Essential Links
- [GitHub Repo for Developing Data Products](https://github.com/yohance-nicholas/Developing-Data-Products)
- [Shiny Application](https://yohance-nicholas.shinyapps.io/covid19_regression_map/)
- [Project overview and instructions](https://www.coursera.org/learn/data-products/peer/tMYrn/course-project-shiny-application-and-reproducible-pitch)


## Mapping the Pandemic with Leaflet() 

```{r message=FALSE, warning=FALSE}
source("https://git.io/JfYVv", local = T)
library(leaflet)
library(leaflet.extras)
caricom_covid_map <- leaflet() %>% 
  addProviderTiles(providers$CartoDB.DarkMatter) %>% 
  addMarkers(lat = caricom_today$lat, 
             lng = caricom_today$lng,
             popupOptions = markerClusterOptions,
             popup = paste("<b>", caricom_today$country, "</b> <br>",
                           "Date: ", caricom_today$date, "<br>",
                           "Confirmed Cases: ", caricom_today$confirmed, "<br>",
                           "Deaths:", caricom_today$deaths,"<br>",
                           "Recovered", caricom_today$recovered,"<br>",
                           "Income Group:", caricom_today$income, "<br>",
                           "Population Aged 65+:", caricom_today$pop_65_over_2018,"<br>",
                           "Diabetes prev. (% pop 20-79):", caricom_today$diabetes_20_79)) %>% 
  addCircleMarkers(lat = caricom_today$lat, 
                   lng = caricom_today$lng,
                   weight = 1,
                   radius = caricom_today$confirmed_per_100k,
                   color = 'yellow') %>% 
  addCircleMarkers(lat = caricom_today$lat, 
                   lng = caricom_today$lng,
                   weight = 1,
                   radius = caricom_today$deaths_per_100k,
                   color = 'red') %>%
  addLegend("bottomright", 
            colors= c('red', 'yellow'), 
            labels= c('Deaths per 100,000', 'Confirmed Cases per 100,000'), 
            title="Legend") 
caricom_covid_map 
```
